Comment on Hodgkins et al. (2018)

### 2023-11-27 analysis/paper/002-reply-main.Rmd
Update the comment on Hodgkins et al. (2018) with comments from Jeff Chanton (received via email, 2023-11-26).

That said, theoretical considerations of decomposition processes alone support the suggested gradient in peat chemistry and @Hodgkins.2018 made an important contribution in providing the first open access models and estimates for peat carbohydrate and aromatics content on a large scale. We hope that our comment highlights that accuracy matters and that it helps to focus research on the problems which hamper improving this accuracy.

@@

Consequently, in general --- and especially in (sub)tropical peatlands --- the average residual enrichment of Klason lignin due to decomposition probably has been underestimated by the original model. A consequence of this in light of climate change is that high latitude peat deposits may be even more prone to shifts in decomposition rates than suggested in @Hodgkins.2018.

@@

However, using precise concepts for organic matter fractions is also important for exactly the opposite reason: to not suggest a variable has more precision than is actually the case; this becomes especially important when interpreting peat chemistry based on mid infrared spectra (MIRS) or spectral prediction models because all variables derived from MIRS --- for example the widely used humification indices, as proxy for relative contents of recalcitrant organic matter fractions (e.g. @Broder.2012), or Klason lignin contents predicted from models --- are in fact only indirect estimates of organic matter fractions and there are abundant possibilities how they can be misleading if they are not sufficiently validated. Stating these limitations requires precise concepts.

Intensities recorded in MIRS really only represent the fraction of the incident infrared radiation which is absorbed by specific molecular structures which happen to absorb at that specific energy level [@Stuart.2004]. Such molecular structures may for example be aromatic C=C bonds, C-N bonds in amides, or C-O bonds in alcohols (e.g. carbohydrates) [@Stuart.2004]. In addition, absorption of infrared radiation (and hence peak heights or areas) can be decoupled from abundances of molecular structures because also the electronegativity of the bonds controls absorption of mid infrared radiation [@Stuart.2004]. These confounding factors make it necessary to explicitly define the conditions under which a spectral prediction model or other variable derived from MIRS (e.g. a humification index) is a valid proxy for a specific, precisely defined, organic matter fraction.

One major reason why the model for Klason lignin in @Hodgkins.2018 is biased and invalid for peat is that the only predictor variable used in this model, `arom15arom16`, is not a good indicator for Klason lignin because also proteins can absorb in the same energy range [@Stuart.2004] and therefore, estimates are biased depending on the amount of proteins in the samples [@Teickner.2022f]. Therefore, we should interpret `arom15arom16` and similar variables, such as specific humification indices (HI$_{1630/1090}$, HI$_{1510/1090}$) only then as good proxies for the relative abundance of aromatic C=C bonds if we have shown that protein contents do not differ much between the peat samples (thresholds which still need to be established).

@@

@Hodgkins.2018 have made an important first step to actually quantify holocellulose and Klason lignin contents, i.e. specific concepts of carbohydrates and aromatics, from MIRS and this is an important improvement over the qualitative interpretation of peak heights or peak ratios (such as humification indices) used in the past (e.g. @Cocozza.2003, @Broder.2012, @Tfaily.2014). Using precise words for what we have actually measured and what confounds we have considered is important to assure that this first step will actually be an improvement over the qualitative interpretation of MIRS.

>>>>

That said, theoretical considerations of decomposition processes alone support the suggested gradient in peat chemistry and @Hodgkins.2018 made an important contribution in providing the first open access models and estimates for peat carbohydrate and aromatics content on a large scale. Increasing the accuracy of models is an important part of the scientific process and we hope that this comment will focus research on the problems which hamper improving this accuracy.

@@

Consequently, in general --- and especially in (sub)tropical peatlands --- the average residual enrichment of Klason lignin due to decomposition probably has been underestimated by the original model. A consequence of this is that high latitude peat deposits may experience more decomposition under a warmer climate than suggested in @Hodgkins.2018 because a larger content of Klason lignin is necessary to stabilize peat chemically under warmer conditions.

@@

However, using precise concepts for organic matter fractions is also important for exactly the opposite reason: to make clear how much accuracy and precision a variable actually has, instead of implying that it would quantify a variable more accurate and precise than is actually the case. This becomes especially important when interpreting peat chemistry based on mid infrared spectra (MIRS) or spectral prediction models because all variables derived from MIRS --- for example the widely used humification indices, as proxy for relative contents of recalcitrant organic matter fractions (e.g. @Broder.2012), or Klason lignin contents predicted from models --- are in fact only indirect estimates of organic matter fractions. These indirect estimates of organic matter fractions can be misleading in abundant ways if they are not sufficiently validated. Stating these limitations requires precise concepts.

Intensities recorded in MIRS really only represent the fraction of the incident infrared radiation which is absorbed by specific molecular structures which happen to absorb at that specific energy level [@Stuart.2004]. Such molecular structures may for example be aromatic C=C bonds, C-N bonds in amides, or C-O bonds in alcohols (e.g. carbohydrates) [@Stuart.2004]. When molecular structures absorb infrared radiation, this causes a change in dipole moment of one or more of their bonds and the larger the change in dipole moment, the more intense is the absorption and hence the larger the peak in a MIRS. The same stretching results in a larger change in dipole moment for more electronegative bonds than less electronegative bonds [@Stuart.2004]. Since C-N bonds are more electronegative than aromatic C=C bonds, this means that already a small amount of proteins in peat can contribute equally large or more to the peaks in MIRS around 1510 and 1630 cm$^{-1}$ than aromatic C=C bonds in the same region of MIRS (see Fig. 3 in @Reuter.2020 for an example). These confounding factors make it necessary to explicitly define the conditions under which a spectral prediction model or other variable derived from MIRS (e.g. a humification index) is a valid proxy for a specific, precisely defined, organic matter fraction.

One major reason why the model for Klason lignin in @Hodgkins.2018 is biased and invalid for peat is that the only predictor variable used in this model, `arom15arom16`, is not a good indicator for Klason lignin because, as explained above, also proteins can absorb in the same energy range [@Stuart.2004] and therefore, estimates are biased depending on the amount of proteins in the samples [@Teickner.2022f]. Therefore, we should interpret `arom15arom16` and similar variables, such as specific humification indices (HI$_{1630/1090}$, HI$_{1510/1090}$) only then as good proxies for the relative abundance of aromatic C=C bonds if we have shown that protein contents do not differ much between the peat samples (thresholds which still need to be established).

@@

@Hodgkins.2018 have made an important first step to actually quantify holocellulose and Klason lignin contents, i.e. specific concepts of carbohydrates and aromatics, from MIRS and this is an important improvement over the qualitative interpretation of peak heights or peak ratios (such as humification indices) used in the past (e.g. @Cocozza.2003, @Broder.2012, @Tfaily.2014). Using precise words for what we have actually measured and what confounding factors we have considered is important to assure that this first step will actually be an improvement over the qualitative interpretation of MIRS.
